---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r setoptions, echo=FALSE}
opts_chunk$set(echo=TRUE)
```
## Introduction
Dear Peer Reviewer,  What follows is the first course assignment for the Coursera module "Reproducible Research". On occasion, I have gone beyond what is required by the question; I'm aware that plagiarism is a constant worry to us all, and I hope that by adding one or two extra pieces of information I should keep it 'different'. Hopefully, I won't have added significantly to your job in marking.

Thank you for your time.

## Loading and preprocessing the data
The data are stored in a zip file, located in the working directory. Our first step is to extract it.
```{r extract}
data1 <- read.csv(unz("activity.zip","activity.csv"))
```
We need to convert to the correct class of R object, which we can do with:
```{r convert}
data1$steps    <- as.numeric(data1$steps)
data1$date     <- as.Date(data1$date, "%Y-%m-%d")
data1$interval <- factor(data1$interval)
```

## What is mean total number of steps taken per day?
Here I've used the dplyr package to group by date and summarize.
```{r groupandsummarize}
library(dplyr)

data2 <- tbl_df(data1)  %>%
        group_by(date)  %>%
        summarise(total = sum(steps,na.rm=TRUE))
```
To show this graphically, I've used the ggplot2 package.
```{r hist}
library(ggplot2)
```

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
